Molecular dynamics simulation of synchronization of a driven particle

### 2021-04-11 manuscript/manuscript_ajp_version3.tex
rearranging introduction

%DM replace?
%\title{Molecular dynamics simulation of synchronization in driven particles}
\title{Molecular dynamics simulation of a driven synchronized particle}

\author{Tiare Guerrero}
\email{guer9330@pacificu.edu}
\affiliation{Department of Physics, Pacific University, Forest Grove, OR 97116}

\author{Danielle McDermott}
\email{mcdermott@pacific.edu}
\affiliation{Department of Physics, Pacific University, Forest Grove, OR 97116}

@@

\begin{abstract}
  Synchronization
  plays an important role in many physical processes.
  We discuss a 
  molecular dynamics simulation
  of a single particle
  moving through
  a viscous liquid
  while being driven 
  across a washboard potential energy landscape.
  Our results show many dynamical patterns
  as the landscape and driving force are altered.
  For certain conditions,
  the particle velocity and location
  are synchronized or 
  phase-locked,
  forming closed orbits in phase space.
  Quasi-periodic motion is common, 
  for which the
  dynamical center of motion shifts the
  phase space orbit.
  This synchronized motion
  is
  observed in 
  simulations and table-top experiments,
  and
  can be used to isolate
  complex natural behaviors.
 %[suggest omitting this sentence. Code not in text. We include molecular dynamics code
 % to simulate and characterize the
  %particle dynamics.] 
\end{abstract}

@@

\section{Introduction} 
Synchronization is a universal phenomena
in which
individual oscillators change frequency due
to external stimuli.\cite{Pikovsky2003}
The
flickering patterns of
candle flames [DM - the air molecules are oscillating xx is a flame an oscillator? xx]
mediated by temperature fluctuations,\cite{Okamoto2016}
the vibrations of singing wine glasses interacting 
through sound waves,\cite{Arane2009}
and metronomes vibrating through a supporting platform\cite{Jia2015}
are examples of in-phase coupled oscillations. 
Biological systems benefit from cooperative
synchronization --
birds coordinate wing flaps
to optimize energy use during flight,\cite{Portugal2014}
frogs alternate croaking patterns,\cite{Aihara2014}
humans clap in time with music,\cite{Tranchant2016}
and 
at a cellular level, 
neurons simultaneously fire in cardiac muscle\cite{MartinHall1999}
and brain tissue.\cite{Singer1999}
External forcing can cause or regulate 
synchronization. For example, 
an electrical pacemaker 
regulates a heart beat 
and 
a pulsed light modifies the
flashing pattern of fireflies.

Synchronized phase-locking or mode-locking 
first appeared in the scientific literature
with 
Huygens' 1665 experiments on the
motions of synchronized 
pendula in wall-mounted clocks.\cite{Bennett2002}
A locked-mode is an integer frequency ratio.
In Huygens clocks,
the pendula were observed to swing at the  same rate
in the same direction (1:1 mode) or
opposite directions ($-1$:1 mode).
A 2:1 mode occurs when a simple pendulum four
times the length of another swings with twice the period.

Complex dynamics such as synchronized mode-locking
can be studied with
colloid particles in experiments.
Typical colloids are   
plastic spheres suspended in
de-ionized water or silica beads suspended in organic solvent.
Because 
colloids are large 
and move slowly, the
particle positions 
can be measured in real time 
with an optical camera.\cite{Pertsinidis2001}
Experimental measurements of the
step-by-step dynamics of 
colloids performing phase-locked motions
are useful for
understanding synchronization at a single particle level.\cite{Juniper2015,Juniper2017}

@@

The model discussed in this paper
resembles an overdamped driven pendulum.
A single particle oscillator
in a potential well is 
like a skateboarder in a half-pipe or
a child on a swing.
A confined oscillator may 
synchronize its location
to the periodic pattern of the external drive,
moving back and forth in time with
the beat,
or moving between substrate minima.
When 
a constant or 
dc drive is applied,
the particle velocity is modulated by 
the potential energy landscape exerted on the particle.  %seen by the particle.
[DM swapped ``seen by'' for ``exerted on'' xx check added wording xx] 
Below some threshold  
the dc force is not strong enough to push the particle
across a potential maximum so the average velocity is zero,
a phenomena referred to as pinning.\cite{Reichhardt2017}
Above the pinning threshold,
a particle subject to a constant
drive force  increases its speed at a rate proportional
to the external drive.  
When the applied force varies periodically,
the ac drive can 
cause the particle to hop back and forth across
the landscape minima.
Many synchronized patterns occur,
controlled by the substrate period,
leading to 
mode-locking,
where the average particle velocity
is fixed for a range of dc drive forces.\cite{Reichhardt2015}

In this paper we discuss numerical simulations 
of the synchronized dynamics
of a confined particle driven over
a washboard shaped potential energy landscape.
We describe
our molecular dynamics model  in Sec.~\ref{sec:MD}.
The model
is easy to simulate yet relevant
to condensed matter systems.
We summarize our results 
including the synchronized motion of a single confined particle
driven across a periodic landscape in 
Sec.~\ref{sec:results}.

In Sec.~\ref{sec:conclusion}
we describe how our results apply 
to physical systems,
such as optically trapped colloids
%charge density waves in dusty plasma
[DM rewrote to better align with conclusion xx added words is this correct? xx]
and superconducting vortices in the presence of periodic pinning arrays.
% and Josephson junctions.
[DM resolved xx Josephson junctions not explicitly mentioned in conclusion, xx] 
We include problems for interested readers 
in Sec.~\ref{sec:problems}.

\section{Simulation}
\label{sec:MD}
We use a classical model for 
studying the dynamics, %xx you only use single particles xx]
using the net force on a particle to calculate
its trajectory.
The particle is confined in a two-dimensional (2D) [DM singular]
simulation of area $A = L \times L$, where $L=46.\bar{6}\,a_0$
where $a_0$ is a dimensionless unit of length.
[DM - yes.  in the simulation we just use $L=46.6$ and the $a_0$ is to satisfy readers unfamiliar with simulation units xx confused about the role of $a_0$ because it and hence $L$ are dimensionless, Is it really the size of a colloid particle, which in the simulation you set to unity?  xx]
The particle %Particle $i$
[DM resolved xx since we are doing non-interacting particles and thus only one particle is moving can we eliminate the subscript $i$? xx]
has
position $\vec{r} = x \hat{x} + y \hat{y}$
and velocity $\vec{v} = d\vec{r}/dt$.
The edges of the system are treated by
periodic boundary conditions,
such that a particle leaving the edge of the system is mapped
back to a position within the simulation boundaries 
by the transformation $x+L \rightarrow x$ and $y+L \rightarrow y$.
We show a schematic of the system in Fig.~\ref{fig:1_landscape}(a).
The units of the simulated variables are summarized in Table~\ref{tab:1}.
%\footnote{To learn more about the effects of temperature on this simulation see Problem~\ref{ex:brownian}, where we study $U_0 \sim k_B T$.}

We confine the particles using a position dependent 
potential energy, called a landscape or substrate.
The landscape is modulated in the $y$-direction
by the periodic function 
 \begin{equation}
   U(y) = U_0 \cos{(2 \pi y / \lambda)},
     \label{eq:ysubstrate}
\end{equation}
where $\lambda=L/N_p$, with $N_p$ equal to the number of periods,
and $U_0$ is a parameter
 that sets the depth of the minima
 with simulation units of energy $E_0$. 
 We plot this function in 
 Fig.~\ref{fig:1_landscape}
 for $N_p = 3$.  In Fig.~\ref{fig:1_landscape}(a) we show 
 the $x$-$y$ plane with a contour plot of $U(y)$ 
 to illustrate
 the 2D potential energy landscape;
 the maxima are colored gray and the minima colored white.
 [DM - switch from red/blue to gray/white - xx make sure that all your figures make sense when printed in b/w xx]

The confining force on the particle [DM removed subscripts] $i$
 is given by
 \begin{equation}
 \vec{F}_{\ell}(\vec{r}) = - \vec \nabla U(\vec{r}). [DM removed ^{\ell}]
 \label{eq:dudr}
 \end{equation}
 %[xx what does the superscript $\ell$ mean? xx]
 where $\ell$ denotes the landscape.
 In Fig.~\ref{fig:1_landscape}(b) we plot  
 $U(y)$ to illustrate how the magnitude
 $|\vec{F}_{\ell}|$ is calculated from the particle position $y$.
 [DM - begrudingly changed - My collaborators and I usually use superscripts for these equations to separate the force superscripts from the particle index subscripts.  This is consistent with the textbook I teach from.  xx suggest that you use subscripts instead of superscripts. We can easily make that change in the next iteration of your manuscript. However, you would need to change the figures, e.g., $\vec{F}_{\rm d}$ $F_{\rm dc}$ $F_{\rm ac}$. One of us didn't like superscripts where he was an undergraduate!  xx]

Particles are subject to an external time-dependent driving force
$\vec{F}_{\rm d}(t)$
applied parallel to the $y$-direction.
We model this force as
\begin{equation}
  \vec{F}_{\rm d}(t) = [F_{\rm dc} + F_{\rm ac} \sin(\omega t)] \hat{y},
    \label{eq:drive}
\end{equation}
with 
a constant component $F_{\rm dc}$
and a time dependent component with amplitude $F_{\rm ac}$
with angular frequency $\omega = 2 \pi f$.

The inertia of 
small particles is reduced by interactions
with the fluid particles.\cite{Purcell1977}
We assume 
colloids are
overdamped
so the particles do not accelerate,
that is, they are suspended in a continuous viscous fluid
that dissipates energy supplied externally. 
Newton's second law for an individual particle
is simplified
by the assumption that $\vec{a}$ is zero. 
The overdamped equation of motion for
the velocity $\vec{v}$ of 
an isolated particle is
\begin{equation}
  \eta \vec{v} = \vec{F}_{\ell}(\vec{r}) + \vec{F}_{\rm d}(t).
    \label{eq:motion}
\end{equation}
with friction coefficient $\eta = 1$ in units of $v_0 / F_0$.
%[xx added $v_0$ to table, please check xx] 
The term $-\eta \vec{v}$
is a drag force, which models
energy dissipation due to the fluid. 
We discuss  models for
spheres moving through fluids in 
Problem~\ref{ex:reynolds}.

The equation of motion provides a direct calculation of the velocity
of an individual particle at position $\vec{r}$ 
and simulation time $t$.
The   simulation is controlled by a  loop
which runs to a maximum integer time step.
At each time step
we evaluate the net force on each particle as a function of its position
$\vec{r}(t)$
and then integrate
the equation of motion to move particles
to an updated position.
Because the acceleration is zero,
the integration of the equation of motion
is performed via 
the simple first-order Euler method 
\begin{equation}
  \vec{r}(t+\Delta t) = \vec{v}(t) \Delta t + \vec{r}(t)
    \label{eq:euler}
\end{equation}
for a time step $\Delta t = 0.1\,\tau$.
In 
Problem~\ref{ex:euler}
we describe 
the numerical methods for 
solving differential equations.
%------------------------------------------------------------------
\begin{figure} % [h!]
\centering
\includegraphics[width=\columnwidth]{fig1_landscape.pdf}
\caption{
Schematic of the simulation of a single particle
  driven across a washboard potential energy landscape.
  (a) View of the $x$-$y$ plane. 
  The time-dependent applied driving force $\vec{F}_{\rm d}$
  is parallel to the $y$-axis.
  The landscape is shown with
  [DM changed color scheme from red/blue to gray/white for better BW conversion]
  maxima in the potential energy are gray %#red
  and minima are white %.
  A particle is 
  subject to competing forces of the landscape and applied driving force.
  (b) The potential energy function
  along the $y$-axis $U(y)$.  [DM removed $_{\ell}$]
  The particle in (a) is shown at the same $y$-position.
  The slope of $U(y)$ [DM removed $_{\ell}$ use the subscripts with the potential, but will move to superscript xx used subscript here xx]
  is the 
  magnitude of force $\vec{F}_{\ell}$. 
  }
\label{fig:1_landscape}
\end{figure}
%
\section{Mode-locking of a single particle}
% In subsection headings, only the first word is capitalized.
\label{sec:results}
Here we drive 
a single particle 
across the landscape. 
The numerical implementation of the landscape 
is calculated with Eqs.~\eqref{eq:ysubstrate} and \eqref{eq:dudr}  [DM - right, two negative signs, one from the derivative, the other from the definition of force.  I found the ``mistakenly'' dropped minus sign in the depths of january and added it.  xx but you should get a + sign not a minus sign xx] 
\begin{equation}
  \label{eq:force}
  {F_{\ell}}_y(y) = A_{p} \sin{(2 \pi y / \lambda)},
\end{equation}
where the force is scaled by the parameter $A_{p} = 2\pi U_0/\lambda$.
In this section we fix the landscape parameters
to $A_{p} = 0.1\,F_0$ 
with $N_p=20$ minima, 
corresponding to a spatial period $\lambda = 2.3\,a_0$.
The competition between the driving force and the landscape potential
can produce a variety of hopping patterns in the particle motion. 
The relative values of $F_{\rm ac}$, $F_{\rm dc}$ and $A_p$
control the rate and distance a  particle moves 
forward and backward in the landscape.
If $F_{\rm d}(t) > A_p$, a particle can 
overcome the barrier height of the landscape,
and 
the particle hops between minima in the energy landscape.
If the driving frequency is low,
as in Fig.~\ref{fig:2_Fd_vy_time},  
the driven particle 
moves 
in a pattern 
with the same frequency 
as the time-dependent force $F_{\rm d}(t)$,
but is modulated by the landscape period.
We explore changes in the frequency and different values of $F_{\rm ac}$ 
in Problem~\ref{ex:parameters}.
Here we vary $F_{\rm dc}$ 
while holding the remaining parameters fixed.

%xx AJP does not use footnotes
\begin{table}[h!]
\centering
\caption{Simulation parameters and units with comparable
  experimental values.~\cite{Juniper2015,Juniper2017} The substrate is scaled by our force units, while an experimental landscape is scaled by the Brownian motion of the particles.  }
\begin{ruledtabular}
\begin{tabular}{c c c } 
Quantity & Simulation Units & Experimental values\\
\hline
length &  $a_0 = 1$ & $ a_0 \sim 1.5 \mu m$\\
energy & $E_0 = 1$ & \\ %$ < E_0 <$ \\
%electric potential & $V(r_{ij}) = E_0/r_{ij}$ \\
%energy & $E_0 = q^2{Z^*}^2/4\pi \epsilon \epsilon_0 a_0$ \\
%dimensionless interaction strength & $q$ \\
%effective colloidal charge & $Z^*$ \\
%solvent dielectric constant & $\epsilon \epsilon_0$\\
force & $F_0 = E_0 / a_0$ & \\ %$<F_0<$\\
time &  $ \tau = \eta a_0 / F_0 $ & $ \tau \sim 3$\,s\\
velocity &  $ v_0 = a_0 / \tau $ &  $v \sim 5\,\mu$m/s \\
%driving period & $T = 100 \tau$ & $T \sim 1.3$ s \\ %frequency 0.75 Hz
%juniper uses water–ethanol mixture
substrate period & $\lambda = 2.3\,a_0$ & $\lambda = 3.5\,\mu$m \\
substrate amplitude & $A_p = 0.1\,F_0$ & $U_0 = 25 k_B T \sim 1\, {\rm J}$ \\ %$U_0 = 0.037 E_0$ 
temperature & $T_0 = 0$  & $T \sim 290\,$K  \\
\end{tabular}
\end{ruledtabular}
\label{tab:1}
\end{table}

In Fig.~\ref{fig:2_Fd_vy_time}(a)
we plot $F_{\rm d}(t)$ 
as a function of time with the
constants $F_{\rm dc}=0.07$, 
$F_{\rm ac}=0.07$ and $f=0.01$ cycles per time unit $\tau$.
The temporal period of the driving force is
$T = 1/f = 100\,\tau$.
In Fig.~\ref{fig:2_Fd_vy_time}(b) 
we show the $y$-position of the particle
as a function of time,
where we have
normalized $y$ by $\lambda$.
The initial particle position is $y = 0$. 
The particle moves
in the positive $y$-direction
through a distance $\Delta y = \lambda$  in a time $T$,
so that  
the average velocity 
$\langle {v}_y \rangle = \lambda f$. 
The inset of Fig.~\ref{fig:2_Fd_vy_time}(b)
shows $y$ 
over one period $100\,\tau < t < 200\,\tau$
with 
the contour plot described
in Fig.~\ref{fig:1_landscape}(a).
The motion is synchronized such that the 
driving force is a maximum when the landscape 
force is minimum,
as shown by the coincidence of the
steep slope of 
$y/\lambda$ 
and the maxima in $F_{\rm d}(t)$.
When $F_{\rm d}(t)$ is large,
the particle moves across the substrate maxima,
[DM - color change]
shown in gray with the contour plot.
%DM commented out - too much detail
%The slope of $y/\lambda$ equals zero twice
%during a driving period,
%indicating zero
%forward motion 
%due to a coincidence of the negative
%driving force and motion over the landscape maxima.
%------------------------------------------------------------------------
\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{fig2_Fd_vy_time.pdf}
\caption{(a) The applied driving force $F_{\rm d}(t)$ 
  with parameters $F_{\rm dc} = 0.07$, $F_{\rm ac} = 0.07$ and $f=0.01$.
  (b) 
  The $y$-position of the driven particle
  normalized by the period of the substrate $\lambda$.
  The substrate strength is $A_p=0.1$.
  The inset   shows
  the $y$-position
  through the second period $100\tau<t<200\tau$
 along with the contour plot depicting
  the landscape potential described in Fig.~\ref{fig:1_landscape}(a).
  }
\label{fig:2_Fd_vy_time}
\end{figure}

>>>>

\title{Molecular dynamics simulation of synchronization of a  driven particle}

\author{Tiare Guerrero}
\email{guer9330@pacificu.edu}
\affiliation{Department of Physics, Pacific University, Forest Grove, Oregon 97116}

\author{Danielle McDermott}
\email{mcdermott@pacific.edu}
\affiliation{Department of Physics, Pacific University, Forest Grove, Oregon 97116}

@@

\begin{abstract}
  We discuss a 
  molecular dynamics simulation
  of a single particle
  moving through
  a viscous liquid
  while being driven 
  across a washboard potential energy landscape.
  Our results show many dynamical patterns
  as the landscape and driving force are altered.
  For certain conditions,
  the particle velocity and location
  are synchronized or 
  phase-locked,
  forming closed orbits in phase space.
  Quasi-periodic motion is common, 
  for which the
  dynamical center of motion shifts the
  phase space orbit.
  %DM removed first sentence,
  %which was strong and clear, but isolated in the abstract
  %put those ideas in this final sentence, making it a mouthful [26 words]
  Isolating
  synchronized motion
  in 
  simulations and table-top experiments,
  can be used to study %DM replaced isolate with study
  complex natural behaviors
  which
  play an important role in many physical processes.
 %[suggest omitting this sentence. Code not in text. We include molecular dynamics code
 % to simulate and characterize the
  %particle dynamics.] 
\end{abstract}

@@

\section{Introduction} 
Synchronization is a universal phenomena
in which
individual oscillators change frequency due
to external stimuli.\cite{Pikovsky2003}
The
flickering patterns of
candle flames 
mediated by temperature fluctuations,\cite{Okamoto2016}
the vibrations of singing wine glasses interacting 
through sound waves,\cite{Arane2009}
and metronomes vibrating through a supporting platform\cite{Jia2015}
are examples of in-phase coupled oscillations. 
Biological systems benefit from cooperative
synchronization --
birds  coordinate their wing flaps
to optimize energy use during flight,\cite{Portugal2014}
frogs alternate croaking patterns,\cite{Aihara2014}
humans clap in time with music,\cite{Tranchant2016}
and 
at a cellular level, 
neurons simultaneously fire in cardiac muscle\cite{MartinHall1999}
and brain tissue.\cite{Singer1999}
%DM ended paragraph with fewer examples, added transition sentence
Mechanisms of synchronization
can be WHAT

Synchronized phase-locking or mode-locking 
first appeared in the scientific literature
in 1665 with 
Huygens'  experiments on the
motions of synchronized 
pendula in wall-mounted clocks.\cite{Bennett2002}
A locked-mode is an integer frequency ratio.
In Huygens clocks,
the pendula were observed to swing at the same rate
in the same direction (1:1 mode) or
opposite directions ($-1$:1 mode).
A 2:1 mode occurs when a simple pendulum four
times the length of another swings with twice the period.
%DM transition sentence
Simulations and laboratory experiments
of mode-locked synchronization
can help provide insights to a diverse range of 
oscillation phenomena.

%DM better main topic
Experiments with colloids

Complex dynamics such as synchronized mode-locking
can be studied with
colloid particles.
In experiments %DM moved the ``in experiments'' to second sentence
typical colloids are   
plastic spheres suspended in
de-ionized water or silica beads suspended in organic solvent.
Because 
colloids are large 
and move slowly, the
particle positions 
can be measured in real time 
with an optical camera.\cite{Pertsinidis2001}
Experimental measurements of the
step-by-step dynamics of 
colloids performing phase-locked motions
are useful for
understanding synchronization
at a single particle level.\cite{Juniper2015,Juniper2017}

@@

%new topic - DM this sentence appeared toward the end of the intro
The model discussed in this paper
resembles a damped driven pendulum 
where external forces
add or remove energy from the system.
A single pendulum is driven by
the external force of gravity to return
to an equilibrium position,
and may be damped to decreasing amplitude
by mechanical friction or air resistance.
Multiple pendula may swing in phase
when coupled through shared vibrations.

External forcing
can cause or regulate synchronization.
For example, 
an electrical pacemaker 
regulates a heart beat 
and 
a pulsed light modifies the
flashing pattern of fireflies. 
%damping
%transition

A single particle oscillator
in a potential well is 
like a skateboarder in a half-pipe or
a child on a swing.
A confined oscillator may 
synchronize its location
to the periodic pattern of the external drive,
moving back and forth in time with
the beat,
or moving between substrate minima.
When 
a constant or 
dc drive is applied,
the particle velocity is modulated by 
the potential energy landscape exerted on the particle.   
Below some threshold  
the dc force is not strong enough to push the particle
across a potential maximum so the average velocity is zero,
a phenomena referred to as pinning.\cite{Reichhardt2017}
Above the pinning threshold,
a particle subject to a constant
drive force  increases its speed at a rate proportional
to the external drive.  
When the applied force varies periodically,
the ac drive can 
cause the particle to hop back and forth across
the landscape minima.
Many synchronized patterns occur,
controlled by the substrate period,
leading to 
mode-locking,
where the average particle velocity
is fixed for a range of dc drive forces.\cite{Reichhardt2015}

In this paper we discuss numerical simulations 
of the synchronized dynamics
of a confined particle driven over
a washboard shaped potential energy landscape.
We describe
our   model  in Sec.~\ref{sec:MD}.
The model
is easy to simulate yet relevant
to condensed matter systems.
We summarize our results 
including the synchronized motion of a single confined particle
driven across a periodic landscape in 
Sec.~\ref{sec:results}. In Sec.~\ref{sec:conclusion}
we describe how our results apply 
to physical systems,
such as optically trapped colloids
and superconducting vortices in the presence of periodic pinning arrays.
We include problems for interested readers 
in Sec.~\ref{sec:problems}.

\section{Simulation}
\label{sec:MD}
We use a classical model for 
studying the dynamics, 
using the net force on a particle to calculate
its trajectory.
The particle is confined in a two-dimensional 
simulation of area $A = L \times L$, where $L=46.\bar{6}\,a_0$
and $a_0$ is a dimensionless unit of length. (In the simulations we   set $a_0 = 1$.)
The particle 
has
position $\vec{r} = x \hat{x} + y \hat{y}$
and velocity $\vec{v} = d\vec{r}/dt$.
The edges of the system are treated by
periodic boundary conditions,
such that a particle leaving the edge of the system is mapped
back to a position within the simulation boundaries 
by the transformation $x+L \rightarrow x$ and $y+L \rightarrow y$.
We show a schematic of the system in Fig.~\ref{fig:1_landscape}(a).
The units of the simulated variables are summarized in Table~\ref{tab:1}.

We confine the particles using a position dependent 
potential energy, called a landscape or substrate.
The landscape is modulated in the $y$-direction
by the periodic function 
 \begin{equation}
   U(y) = U_0 \cos{(2 \pi y / \lambda)},
     \label{eq:ysubstrate}
\end{equation}
where $\lambda=L/N_p$, with $N_p$ equal to the number of periods,
and $U_0$ is a parameter
 that sets the depth of the minima
 with simulation units of energy $E_0$. 
 We plot this function in 
 Fig.~\ref{fig:1_landscape}
 for $N_p = 3$.  In Fig.~\ref{fig:1_landscape}(a) we show 
 the $x$-$y$ plane with a contour plot of $U(y)$ 
 to illustrate
 the 2D potential energy landscape;
 the maxima are colored gray and the minima colored white.

The confining force on the particle $i$
 is given by
 \begin{equation}
 \vec{F}_{\ell}(\vec{r}) = - \vec \nabla U(\vec{r}),
 \label{eq:dudr}
 \end{equation}
where $\ell$ denotes the landscape.
 In Fig.~\ref{fig:1_landscape}(b) we plot  
 $U(y)$ to illustrate how the magnitude
 $|\vec{F}_{\ell}|$ is calculated from the particle position $y$.

The particle is subject to an external time-dependent driving force
$\vec{F}_{\rm d}(t)$
applied parallel to the $y$-direction.
We model this force as
\begin{equation}
  \vec{F}_{\rm d}(t) = \big [F_{\rm dc} + F_{\rm ac} \sin(\omega t)\big] \hat{y},
    \label{eq:drive}
\end{equation}
with 
a constant component $F_{\rm dc}$
and a time dependent component with amplitude $F_{\rm ac}$
and angular frequency $\omega = 2 \pi f$.

The inertia of the 
small particle is reduced by interactions
with the fluid particles.\cite{Purcell1977}
We assume 
colloids are
overdamped
so the particle does not accelerate,
that is, it is suspended in a continuous viscous fluid
that dissipates energy supplied externally. 
Newton's second law for the particle
is simplified
by the assumption that $\vec{a}$ is zero. 
The overdamped equation of motion for
the velocity $\vec{v}$ of 
an isolated particle is
\begin{equation}
  \eta \vec{v} = \vec{F}_{\ell}(\vec{r}) + \vec{F}_{\rm d}(t),
    \label{eq:motion}
\end{equation}
with friction coefficient $\eta = 1$ in units of $v_0 / F_0$.
The term $-\eta \vec{v}$
is a drag force that models
energy dissipation due to the fluid. 
We discuss  models for
a sphere moving through a fluid in 
Problem~\ref{ex:reynolds}.

The equation of motion provides a direct calculation of the velocity
of an individual particle at position $\vec{r}$ 
and simulation time $t$. At each time step
we evaluate the net force on the particle as a function of its position
$\vec{r}(t)$
and then integrate
the equation of motion to move the particle
to its updated position.
Because the acceleration is zero,
the integration of the equation of motion
is performed using 
the simple first-order Euler method 
\begin{equation}
  \vec{r}(t+\Delta t) = \vec{r}(t) + \vec{v}(t) \Delta t
    \label{eq:euler}
\end{equation}
for a time step $\Delta t = 0.1\,\tau$.
In 
Problem~\ref{ex:euler}
we describe 
the numerical methods for 
solving differential equations such as Eq.~\eqref{eq:euler}.
%------------------------------------------------------------------
\begin{figure} % [h!]
\centering
\includegraphics[width=\columnwidth]{fig1_landscape.pdf}
\caption{
Schematic of the simulation of a single particle
  driven across a washboard potential energy landscape.
  (a) View of the $x$-$y$ plane. 
  The time-dependent applied driving force $\vec{F}_{\rm d}$
  is parallel to the $y$-axis.
  The landscape is shown with
  maxima of the potential energy in gray 
  and minima in white. 
  The particle is 
  subject to competing forces of the landscape and the applied driving force.
  (b) The potential energy function $U(y)$. 
  The particle in (a) is shown at the same $y$-position in (b).
  The slope of $U(y)$ is the 
  magnitude of the force $\vec{F}_{\ell}$. 
  }
\label{fig:1_landscape}
\end{figure}

\section{Mode-locking of a single particle}

\label{sec:results}
Here we drive 
a single particle 
across the landscape. 
The numerical implementation of the force due to the landscape 
is calculated using Eqs.~\eqref{eq:ysubstrate} and \eqref{eq:dudr}  
\begin{equation}
  \label{eq:force}
  {F_{\ell}}_y(y) = A_{p} \sin{(2 \pi y / \lambda)},
\end{equation}
where the force is scaled by the parameter $A_{p} = 2\pi U_0/\lambda$.
In this section we fix the landscape parameters
to $A_{p} = 0.1\,F_0$ 
with $N_p=20$ minima, 
corresponding to a spatial period $\lambda = 2.3\,a_0$.
The competition between the driving force and the landscape potential
can produce a variety of hopping patterns in the particle motion. 
The relative values of $F_{\rm ac}$, $F_{\rm dc}$, and $A_p$
control the rate and distance a  particle moves 
forward and backward in the landscape.
If $F_{\rm d}(t) > A_p$, a particle can 
overcome the barrier height of the landscape,
and 
the particle hops between minima in the energy landscape.
If the driving frequency is low,
as in Fig.~\ref{fig:2_Fd_vy_time},  
the driven particle 
moves 
in a pattern 
with the same frequency 
as the time-dependent force $F_{\rm d}(t)$,
but is modulated by the landscape period.
We explore changes in the frequency and different values of $F_{\rm ac}$ 
in Problem~\ref{ex:parameters}.
Here we vary $F_{\rm dc}$ 
while holding the remaining parameters fixed.

\begin{table}[h!]
\centering
\caption{Simulation parameters and units with comparable
  experimental values.~\cite{Juniper2015,Juniper2017} The substrate is scaled by our force units, while an experimental landscape is scaled by the Brownian motion of the particles. }%[DM changed to T xx what does $T_0 = 0$ mean?  $T_0$ is never used in the paper. xx]  }
\begin{ruledtabular}
\begin{tabular}{c c c } 
Quantity & Simulation Units & Experimental values\\
\hline
length &  $a_0 = 1$ & $ a_0 \sim 1.5 \mu m$\\
energy & $E_0 = 1$ & \\ %$ < E_0 <$ \\
%electric potential & $V(r_{ij}) = E_0/r_{ij}$ \\
%energy & $E_0 = q^2{Z^*}^2/4\pi \epsilon \epsilon_0 a_0$ \\
%dimensionless interaction strength & $q$ \\
%effective colloidal charge & $Z^*$ \\
%solvent dielectric constant & $\epsilon \epsilon_0$\\
force & $F_0 = E_0 / a_0$ & \\ %$<F_0<$\\
time &  $ \tau = \eta a_0 / F_0 $ & $ \tau \sim 3$\,s\\
velocity &  $ v_0 = a_0 / \tau $ &  $v \sim 5\,\mu$m/s \\
%driving period & $T = 100 \tau$ & $T \sim 1.3$ s \\ %frequency 0.75 Hz
%juniper uses water–ethanol mixture
substrate period & $\lambda = 2.3\,a_0$ & $\lambda = 3.5\,\mu$m \\
substrate amplitude & $A_p = 0.1\,F_0$ & $U_0 = 25 k_B T \sim 1\, {\rm J}$ \\ %$U_0 = 0.037 E_0$ 
temperature & $T = 0$  & $T \sim 290\,$K  \\
\end{tabular}
\end{ruledtabular}
\label{tab:1}
\end{table}

In Fig.~\ref{fig:2_Fd_vy_time}(a)
we plot $F_{\rm d}(t)$ 
as a function of time with the
constants $F_{\rm dc}=0.07$, 
$F_{\rm ac}=0.07$ and $f=0.01$ cycles per time unit $\tau$.
The  temporal period of the driving force is
$T = 1/f = 100\,\tau$.
In Fig.~\ref{fig:2_Fd_vy_time}(b) 
we show the $y$-position of the particle
as a function of time,
where we have
normalized $y$ by $\lambda$.
The initial particle position is $y = 0$. 
The particle moves
in the positive $y$-direction
through a distance $\Delta y = \lambda$  in the time $T$,
so that  
the average velocity 
$\langle {v}_y \rangle = \lambda f$. 
The inset of Fig.~\ref{fig:2_Fd_vy_time}(b)
shows $y$ 
over one period $100\,\tau < t < 200\,\tau$
with 
the contour plot described
in Fig.~\ref{fig:1_landscape}(a).
The motion is synchronized such that the 
driving force is a maximum when the landscape 
force is minimum,
as shown by the coincidence of the
steep slope of 
$y/\lambda$ 
and the maxima in $F_{\rm d}(t)$.
When $F_{\rm d}(t)$ is large,
the particle moves across the substrate maxima.
%shown in gray with the contour plot. said in caption

\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{fig2_Fd_vy_time.pdf}
\caption{(a) The applied driving force $F_{\rm d}(t)$ 
  with parameters $F_{\rm dc} = 0.07$, $F_{\rm ac} = 0.07$ and $f=0.01$.
  (b) 
  The $y$-position of the driven particle
  normalized by the period of the substrate $\lambda$.
  The substrate strength is $A_p=0.1$.
  The inset   shows
  the $y$-position
  through the second period $100\,\tau<t<200\,\tau$
 along with the contour plot depicting
  the landscape potential described in Fig.~\ref{fig:1_landscape}(a).
  }
\label{fig:2_Fd_vy_time}
\end{figure}

To explore the possible hopping patterns,
we sweep through a range of $F_{\rm dc}$ for fixed $F_{\rm ac}$ and $A_p$.
In Fig.~\ref{fig:3_sweep_vyFDC} 
we increase $F_{\rm dc}$ in increments of $0.001\,F_0$,
and 
measure the average velocity $\langle v_y \rangle $ 
as a function of $F_{\rm dc}$.
We also perform the sweep for a non-oscillatory drive $F_{\rm ac} = 0$.
With no oscillating component of the driving force,
  the force-velocity relation increases monotonically 
  above the depinning threshold $F_c$ such that
  \begin{equation}
    \langle v_y \rangle \propto (F_{\rm dc}-F_c)^{-\beta},
  \end{equation}
  where the power $\beta$ varies with the
  %system type and
  type of phase transition and 
  %can
  may 
  be used to identify the universality class.\cite{Reichhardt2017} %[xx what does system type mean? xx - DM modified to phase transition]
  The critical force $F_c$ is equal to the maximum substrate force
  $A_p$.

The addition of an ac drive leads
  to the formation of modes.
  A mode is a periodic pattern of hops
  with a constant average particle velocity, $\langle {v}_{y} \rangle$
  over a range of driving forces $F_{\rm dc}$.
  In Fig.~\ref{fig:3_sweep_vyFDC}
  we sweep $F_{\rm dc}$
  with $F_{\rm ac} = 0.07$ and $f=0.01$.
Each step represents a different pattern of hops
between substrate minima
performed by the particle
due to the landscape confinement.  
At 
low values of $F_{\rm dc}$, the average velocity $\langle v_y \rangle$ is zero.
Because 
$A_p$ is large compared to the extrema of $F_{\rm d}(t)$,
the particle oscillates back and forth
in a single minima with no net velocity, an example of 
a 0:0 mode.
At higher values of $F_{\rm dc}$, the particle velocity 
$\langle v_{y} \rangle$ increases in steps of uniform height,
$\langle v_{y} \rangle = n \lambda f$,
where $n$ is an integer.
We observe a mode of $n=1$
for   $0.05 < F_{\rm dc} < 0.08$,
$n=2$ for $0.08 < F_{\rm dc} < 0.11$,
$n=3$ for $0.12 < F_{\rm dc} < 0.13$,
$n=4$ for $0.14 < F_{\rm dc} < 0.155$, and 
$n=5$ for $0.155 < F_{\rm dc} < 0.16$.
Higher modes are not visible.
The step width is nonlinear
and depends on the strength of $F_{\rm ac}$
for this landscape potential.\cite{Reichhardt2000,Juniper2017}
These steps, known as Shapiro steps, can have a variety of
interesting patterns
such as a devil's staircase related to chaotic dynamics.\cite{Bak1986}
